Natural Selection Across Three Generations of Americans

### 2024-02-20 response-to-reviewers-revision-1.qmd

---
title: "'Natural Selection Across Three Generations of Americans': response to reviewers"
format: docx
editor: 
  markdown: 
    wrap: 72
bibliography: bibliography-hrs.bib
---

@@

David Hugh-Jones

Reviewer Comments:

# Reviewer 1

*Thanks for the opportunity to read the manuscript 'Natural selection in
the Health and Retirement Study'. This study is of great interest to the
behavior genetics and population genetics community and beyond. It
demonstrates ongoing natural selection in a descriptive manner and
partly confirms ideas about the trade-off between childbearing and work.
I have a few comments.*

@@

Note that controlling for principal components in a regression, *y \~
PGIs + PCs*, gives the same effect sizes as running an uncontrolled
regression on PGIs which have previously had the PCs residualized out,
*y \~ PGIs_resid*. We control rather than residualizing because this
gives the correct standard errors[^1], which might matter with our
relatively low N.

[^1]: Because the controls are correctly accounted for when calculating
    degrees of freedom.

Why residualize or control for principal components at all? I think for
two reasons. First, we are ultimately interested in selection on the
"true PGI". Stripping out PCs is an admittedly imperfect way to control
for confounding population stratification. Second, PGIs are typically
residualized in this way, so by examining the effect of these we're
following the "industry standard" and learning about scores in the form
they are actually used.

*In general, the authors state that they are less interested in
significance but in effect sizes. First, this is very generous. Whilst I
find this acceptable, it would be good to read one or two more sentences
on why the authors think this is legitimate in this case. Second, I
believe such a descriptive approach might be useful studying the
patterns of natural selection, however, when it comes to the
interactions, the author state that effects 'Appear larger for
unmarried', for example. It remains unclear to me when an effect is
considered larger or not, which becomes important as the author
differentiates between interactions which are real and those which are
not. What are the criteria here.*

@@

More specifically, we're interested in the pattern of selection effects
across different PGIs, and how it correlates with the PGI's relationship
to income. So we want to include PGIs with small selection effects,
which we predict to have a low absolute correlation to income. Given the
sample size, these PGIs are very likely to have insignificant individual
selection effects, but we shouldn't ignore them in the cross-PGI
analyses.

When we examine relationships across scores, we indeed test for
significance, always by bootstrapping the sample.

*The authors state that the correlation between the PGI and fertility
quantifies natural selection. However, to my knowledge it is the
covariance following the Robert-Price Identity: Morrissey, M. B., L. E.
B. Kruuk, and Alastair J. Wilson. "The danger of applying the breeder's
equation in observational studies of natural populations." Journal of
evolutionary biology 23.11 (2010): 2277-2288.*

We quantify natural selection with the regression coefficient of the PGI
on relative lifetime reproductive success. At times in the original
manuscript this was wrongly referred to as a correlation but has now
been updated. The polygenic scores are standardized to have a standard
deviation of 1. Under a univariate model, the regression coefficient is
the covariance divided by the variance in the independent variable.
Given the variance is 1, the regression coefficient is equal to the
covariance and thus the two are interchangeable.

The story is slightly complicated by the fact that we control for
principal components. Thus our estimates are the partial covariance
divided by the partial variance, the partial variance being slightly
less than 1. Slightly biasing the effect size upwards in magnitude.
Nevertheless, this is the standard approach used by others
[@kong2017selection].

It would be possible to residualise the polygenic score on the principal
components, before restandardizing it have a variance of 1. That we
might think we have something closer to the true polygenic score
unafflicted by population stratification, with the right variance. In
practice we've found this to only change estimates for the effect of the
EA PGI by \~1%. The area of our analysis where the exact estimate is key
is the error-corrected estimates. Since the estimates need to be
adjusted for the fact that the error-corrected polygenic scores will
covary differently with the principal components, making such a similar
adjustment would be rather difficult.

@@

*The authors might want to discuss the expected selection in terms of
relevance. They quantify it but they do not discuss these
quantifications. This is very important shaping our expectation of
ongoing natural selection. It might also be very interesting to know
more about whether those trends follow the secular trends in
phenotypes.*

We have now added a final paragraph to the Discussion section on this.
In particular we discuss PGIs for health phenotypes, as a corrective to
the existing literature which is perhaps a bit over-focused on cognition
and IQ.

TE TODO: Could we look at health trends in phenotypes? I'd guess that
health trends improve over the C20, but we could look at some specifics.
(Depression/BMI?) One big issue here is that maybe the PGIs have
confounds... but perhaps that's less serious for some of these
phenotypes

*Finally, the statement 'in the UK Biobank sample, correlations of PGS
coefficients between younger and older parents are negative (-0.319),
and for 42 out of 58 PGS, coefficients have opposite signs. However, in
bootstraps, these statistics are not significantly different from 0 and
29 respectively.' was not entirely clear to me. What does it mean that
42 PGS had opposite sign effects? Does that mean for old versus younger
parents' coefficients were very different, even in the opposite
direction? How does this relate to the first part of the sentence that
coefficients were negative. Also, why has this been tested against 0 and
29?*

@@

> Whilst past research has focused on \[cognitive\] traits, we find
> selection differentials are often of similar magnitudes for health
> related traits. The selection differential for self-rated health is
> greater than those of education and cognitive performance. The most
> significant, positively selected trait was ADHD, which was also found
> in the UK [@hugh2022human]. Future research should study the health
> and medical implications of natural selection, in addition to their
> social implications.

*What is the data base for the mean number of children of people born in
the same year - is pooling appropriate, as the means may vary
considerably according to ethnicity?*

We thought about this, and now agree that it is better to use mean
number of children of whites and of blacks separately for each ethnic
group:

> "RLRS is calculated separately by ethnicity. This is not ideal,
> because it treats the ethnicities as separate breeding populations.
> But the alternative of calculating pooled RLRS would effectively be
> estimating natural selection in the whole US population by treating
> whites as representative. We therefore focus on the white US
> population, with the caveat that this"one data point" may not be
> representative of the US population (still less the species!) The mean
> number of children of people born in each year was calculated using
> sampling weights."

### 2024-02-20 hrs-selection.qmd
rlrs by race, errors fixed, explained covariance problem, explored top PGI

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for 
  natural selection: scores which correlate negatively (positively) with
  education are selected for (against). The theory predicts that
  selection coefficients should be stronger among low-income, 
  less educated, unmarried and younger parents, but these predictions are only
  partly borne out: coefficients are larger among low-income
  and unmarried parents, but not among younger parents or those with less 
  education. Using respondents' number of siblings and grandchildren, we 
  examine selection across three generations, finding consistent results 
  over time. We also estimate effect sizes corrected for noise in the
  polygenic scores. Selection for some health traits is similar in magnitude
  to that for cognitive traits. 
date: "February 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h13gkid = mean(h13gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h13gkid = h13gkid/mean_h13gkid,
    mean_div_h13gkid = mean(div_h13gkid*child_weight/mean_child_weight, 
                            na.rm = TRUE), 
    rlrs_div_h13gkid = div_h13gkid/mean_div_h13gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

@@

We estimate coefficients of PGS on RLRS. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, and self-rated health. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

>>>>

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for 
  education are selected for (against). The theory predicts that
  selection coefficients should be stronger among low-income, 
  less educated, unmarried and younger parents, but these predictions are only
  and unmarried parents, but not among younger parents or those with less 
  education. Using respondents' number of siblings and grandchildren, we 
  examine selection across three generations, finding consistent results 
  over time. We also estimate effect sizes corrected for noise in the
  polygenic scores. Selection for some health traits is similar in magnitude
  to that for cognitive traits. 
  natural selection: scores which correlate negatively (positively) with
  partly borne out: coefficients are larger among low-income
date: "February 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

rand <- rand |>
  mutate(.by = c(ethnicity, rabyear), # by year and ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h13gkid = mean(h13gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h13gkid = h13gkid/mean_h13gkid,
    mean_div_h13gkid = mean(div_h13gkid*child_weight/mean_child_weight, 
                            na.rm = TRUE), 
    rlrs_div_h13gkid = div_h13gkid/mean_div_h13gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

@@

We estimate coefficients of PGS on RLRS. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf_white` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, and self-rated health. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

```{r}
#| label: Top PGS
#| eval: false

# Is it ok to residualise PGS on PCS vs control for them?
# Frisch-Waugh-Lovell theorem: it's the same thing... But we need PGS standardized to sd of 1
# if the regression beta is to be the covariance for the Price Equation
# When you residualise on PCs *and then restandardize* then the beta *is different* than just controlling for PCs. The difference is marginal (~1% lower in the below example). Furthermore, controlling for PCs seems to have been the standard practice in Beauchamp and in Kong et al.

# I'm not sure whether we can residualise on PCs then use the error correction because PGS true, will correlate differently with the PCs. Then again I'm not sure that the correction would be right for twin H2 since the PCs will surely only correlate with the common SNPs? I've consulted the Becker et al 2021 supplementary and it doesn't give advice on the use of PCs. I'd have to think very carefully about how to do this.

# Either we justify our approach as reasonable. Or we residualise on the PCs and go with it.
# I'm in favour of the latter, in part because explaining why the effect sizes switch sign in the error correction is awkward.

reg_simple <- lm(rlrs ~ scale(EA) ,rand$variables)
reg_pc <- lm(rlrs ~ scale(EA) + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
# beta = -0.027651
reg_res_EA <- lm(`scale(EA)` ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)

reg_res_rlrs <- lm(rlrs ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)

reg_res_noscale <- lm(reg_res_rlrs$residuals ~ reg_res_EA$residuals, 
                               reg_pc$model)
# beta = -0.027651
reg2_res <- lm(reg_res_rlrs$residuals ~ scale(reg_res_EA$residuals),
                       reg_pc$model)
# beta = -0.027358

reg2_res_pgs_only <- lm(rlrs ~ scale(reg_res_EA$residuals),
                       reg_pc$model)
# beta = -0.027358

# Below I get the estimate of the interactions and their p values. 
# EA is only significant for income at p = 0.0106
# There are always traits significant after bonferroni. But usually it's autism
# Given DHJ's improvement to manuscript, I don't think we have anything more convincing to give by focusing on the top PGS.

for (eth in "white") {
  r_eth <- rand |> subset(ethnicity == eth)

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, "*edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_ed[[eth]] <-
    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_ed[[eth]]) <- pgs
  tidy_pgs_x_ed[[eth]] <-
    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
    filter(grepl(':', term)) |>
    mutate(edu = str_remove(term, ".*:"))

  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(grepl(':', term)) |>
    mutate(income_med = str_remove(term, ".*:income_med"))

  form_pgs_x_mar <- paste0(pgs, "*married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(grepl(':', term)) |>
    mutate(
      married = str_remove(term, ".*:married")
    )

  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(grepl(':', term)) |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    ) 
}
